---
title: "P8105 Data Science I Homework 2"
author: Olivia Wang (hw2852)
output: github_document
date: "2022-10-05"
---

# Problem 1 

## 1.1 Read, Clean, Select, Mutate Transit Data

We will first read and clean the CSV data set containing the NYC Transit data. To do so, we will first load in the `tidyverse` package that contains the necessary functions. We will also retain the following columns (i.e., variables) using the `select` function: 

* Line
* Station
* Name
* Station Latitude
* Station Longitude
* Routes Served
* Entry
* Vending
* Entrance Type
* ADA Compliance

```{r}
library(tidyverse)
transit_data = 
  read_csv("./NYC_Transit_Subway_Entrance_And_Exit_Data.csv") %>% 
  janitor::clean_names(.) %>% 
  select(., line, station_name, station_latitude, station_longitude, starts_with("route"), entry, vending, entrance_type, ada)
```

Next, we will convert the `entry` variable from a character (YES vs. NO) to a logical variable using the `ifelse` function. 
```{r}
transit_data = mutate(transit_data, entry = ifelse(entry == "YES", TRUE, FALSE))
```

Let us now generate a summary of the `transit_data` data set, and determine the number of rows and columns in this data set. 
```{r}
summary(transit_data)
nrow(transit_data)
ncol(transit_data)
```

The newly created `transit_data` data set contains __1868 rows__ and __19 columns__.So far, we have loaded in the CSV data set containing NYC Transit data. We then applied the `clean_names` function to clean up variable names. The `select` function was applied to retain the prescribed 19 variables. The `entry` variable was then converted from a character variable (YES vs. NO) to a logical variable (TRUE vs. FALSE). This tidied data set can now be used for further analysis. 


## 1.2 Enumerating Distinct Stations

To determine the number of distinct stations, the `distinct` function can be applied. To do so, we must first load in the `dplyr` package that contains the function. 
```{r}
library(dplyr)
distinct(transit_data, line, station_name, .keep_all = TRUE)
```

Based on the output generated from the code above, we can conclude that there are __465__ distinct stations. 

